---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
library(CorMID)
```


# CorMID

<!-- badges: start -->
[![R-CMD-check](https://github.com/janlisec/CorMID/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/janlisec/CorMID/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![test-coverage](https://github.com/janlisec/CorMID/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/janlisec/CorMID/actions/workflows/test-coverage.yaml)
[![codecov](https://codecov.io/gh/janlisec/CorMID/branch/main/graph/badge.svg?token=NSY6DITZVH)](https://app.codecov.io/gh/janlisec/CorMID)
[![CRAN status](https://www.r-pkg.org/badges/version/CorMID)](https://CRAN.R-project.org/package=CorMID)
[![DOI](https://img.shields.io/badge/doi-10.3390/metabo12050408-yellow.svg)](https://doi.org/10.3390/metabo12050408)
<!-- badges: end -->

**CorMID** is an R-package providing functions to solve problems during 
metabolic flux analysis using HR-APCI-MS.

In metabolic flux experiments tracer molecules (often glucose containing 
labelled carbon) are incorporated in compounds measured using mass spectrometry. 
The mass isotopologue distributions (MIDs) of these compounds needs to be 
corrected for natural abundance of labelled carbon and other effects, which are 
specific on the compound and ionization technique applied. This package provides 
functions to correct such effects in high resolution gas chromatography 
atmospheric pressure chemical ionization mass spectrometry (GC-HR-APCI-MS) 
analyses.

## Installation

You can install the development version of CorMID from 
[GitHub](https://github.com/) with:

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("janlisec/CorMID")
```

or install the version from [CRAN](https://cran.r-project.org/package=CorMID) 
instead.

## Quick Example

**CorMID** is supposed to disentangle a complex MID. Complex means that the ion
intensities of the isotopes are influenced by natural abundance, artificial
labeling (e.g. by a <sup>13</sup>C-Glucose tracer) and mass spectrometry 
artifacts (i.e. several potential adducts).

You can create and visualize such a complex mass spectrum by providing a 
*chemical formula*, the *true MID* and an *adduct distribution* like 
follows:

```{r exmpl1}
# a chemical formula, here: Lactic acid 2 TMS
fml <- "C9H22O3Si2"

# the true mass isotopologue distribution, here: 10% U13C enriched
mid <- c(0.9, 0, 0, 0.1)

# adduct distribution, here: 3 different APCI adducts formed
r <- list("M+H" = 0.8, "M-H" = 0.1, "M+H2O-CH4" = 0.1)

# reconstruct the measured intensity vector
rMID <- CorMID::recMID(mid = mid, r = r, fml = fml)
round(rMID, 3)
```

**CorMID** provides a class specific plotting function for such a reconstructed 
MID:

```{r exmpl1_plot, fig.width=4, fig.height=3.5}
plot(rMID, ylim=c(0,0.6))
mtext(text = "Reconstructed MID", side = 3, line = -1.25, adj = 0.98, outer = T)
text(x = 3, y = 0.4, labels = "[M+H]+", pos=2)
```

Assuming that you have measured these intensities in your experiment, the main 
function of **CorMID** can estimate the underlying *MID* and *r* for you:

```{r exmpl2}
# disentangle the adduct ratios and true isotopologue distribution (enrichment) from the above test data
out <- CorMID::CorMID(int = rMID, fml = fml)
print(out)
```

Please note: no information regarding the true labeling status and the adduct 
distribution was provided in the above function call. **CorMID** is able to 
*guess* the most likely combination.

This allows you to perform the correction for natural abundance and technical 
artifacts in a single step and extract the relevant labeling status for flux
analysis or other statistical evaluations.

## Detailed documentation

You might either read the [Vignette](https://cran.r-project.org/package=CorMID/vignettes/CorMID.html) 
describing the package functions in detail or read the [publication](https://doi.org/10.3390/metabo12050408) 
which shows a evaluation of the performance of **CorMID** on real data sets.
